Tool-driven Revolutions in Archaeological Science

### 2018-09-09 analysis/paper/jcaa-paper.Rmd
added two sentences about what truly would be revolutionary: no black box, evaluation of results, equality in access to research community

These archaeology articles that include publicly available R code and data are part of a new approach to scientific communication that has also been observed in other fields. Although we are not aware of attempts to quantify this change in the same way we have for archaeology, we have observed a number of recent publications that describe and recommend code sharing in statistics [@Baumer2014], genome biology [@markowetz2015five], computational biology [@Sandve2013], biostatistics [@Peng2009], computer science [@peng2011reproducible; @Mitchell2012], applied mathematics [@LeVeque2012], speech science [@Abari2012], political science [@king1995replication; Dafoe2014], and the social sciences generally [@Miguel2014]. As part of this growing interest in using code for research we also see manifestos aimed at researchers doing any kind of quantitative work [@nosek2015promoting; @Ince2012; @Barnes2010] and articles recommending using R in undergraduate education [@Bray2014; @Eglen2009]. Much of this literature is concerned with identifying and solving problems of irreproducibility in research. This comes from a renewed interest in basic principles of the scientific method: that once-off results should not be trusted, instead they should be reproducible by other members of the research community. Furthermore, an article should sufficiently describe the presented research results such that a colleague can fully understand the results and how they were obtained. Many of these authors trace problems of irreproducibility to the increased complextity of computer-based analyses (but only a limited space to describe them in a journal article), and the use of mouse-driven computer programs where the researcher's analytical decisions are not recorded during the research process, and thus cannot be shared with other researchers. In our view, using R (or similar open source programming languages) to conduct research might not constitute a tool-driven revolution, but publicaly sharing the code used for research, as a solution to problems of irreproducibility could enable revolutionary change, since openly shared code could speed the adoption of code throughout a research community.

@@

A pragamtic issue relating to wide-spread changes in scientific practice is what the incentives are for members of the research community that might not be motivated by abstract basic principles of the science method. If code sharing has some concrete career advantages then we might expect it to be more likely to be part of a tool-drive revolution.  @Vandewalle2012 analysed citation rates for 645 articles published in _IEEE Transactions on Image Processing_ during 2004–2006. He found statistically significant differences between the median number of cita?tions for papers that have code available and papers that do not have code available. In Figure \@ref(fig:citation-advantage) we conducted a similar anlysis for archaeology articles. We see that archaeological articles that cite R are consistently more highly cited themselves than articles that do not cite R.

>>>>

These archaeology articles that include publicly available R code and data are part of a new approach to scientific communication that has also been observed in other fields. Although we are not aware of attempts to quantify this change in the same way we have for archaeology, we have observed a number of recent publications that describe and recommend code sharing in statistics [@Baumer2014], genome biology [@markowetz2015five], computational biology [@Sandve2013], biostatistics [@Peng2009], computer science [@peng2011reproducible; @Mitchell2012], applied mathematics [@LeVeque2012], speech science [@Abari2012], political science [@king1995replication; Dafoe2014], and the social sciences generally [@Miguel2014]. As part of this growing interest in using code for research we also see manifestos aimed at researchers doing any kind of quantitative work [@nosek2015promoting; @Ince2012; @Barnes2010] and articles recommending using R in undergraduate education [@Bray2014; @Eglen2009]. Much of this literature is concerned with identifying and solving problems of irreproducibility in research. This comes from a renewed interest in basic principles of the scientific method: that once-off results should not be trusted, instead they should be reproducible by other members of the research community. Furthermore, an article should sufficiently describe the presented research results such that a colleague can fully understand the results and how they were obtained. Many of these authors trace problems of irreproducibility to the increased complextity of computer-based analyses (but only a limited space to describe them in a journal article), and the use of mouse-driven computer programs where the researcher's analytical decisions are not recorded during the research process, and thus cannot be shared with other researchers. In our view, using R (or similar open source programming languages) to conduct research might not constitute a tool-driven revolution, but publicaly sharing the code used for research, as a solution to problems of irreproducibility could enable revolutionary change: Openly shared code could speed the widespread adoption of the true scientific method, since it frees science from the black box, enabling researchers to not only rerun the shared analysis, but to gain access to all parameter settings, empowering them to change these and therefore truly evaluate the published results. As open source languages such as R (or Python...) are free as are the publishing platforms (such as zenodo, github or similar), they also enable not-affluent researchers world wide to contribute equally to the research community.

@@

A pragamtic issue relating to wide-spread changes in scientific practice is what the incentives are for members of the research community that might not be motivated by abstract basic principles of the science method. If code sharing has some concrete career advantages then we might expect it to be more likely to be part of a tool-drive revolution.  @Vandewalle2012 analysed citation rates for 645 articles published in _IEEE Transactions on Image Processing_ during 2004–2006. He found statistically significant differences between the median number of citations for papers that have code available and papers that do not have code available. In Figure \@ref(fig:citation-advantage) we conducted a similar anlysis for archaeology articles. We see that archaeological articles that cite R are consistently more highly cited themselves than articles that do not cite R.
